Probabilistic forecasting of hourly Emergency Department arrivals

### 2022-12-29 paper/paper.Rmd
incorporating changes proposed by Ivan in the response letter

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average
(ARIMA) [@hyndman2021forecasting], Autoregressive Conditional
Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive
model [@lutkepohl2013vector], TBATS [@de2011forecasting] and Artificial
Neural Networks [@hyndman2021forecasting]. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient [@wright2006reexamining]. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently [@ramos2013probabilistic;@ROSTAMITABAR20221197]. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow _optimal_ decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{Another drawback of existing studies is that the
datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
Finally, all previous publications
referenced in this paper are not fully reproducible as underlying data and code are not available.}

@@

# Research background: hourly ED forecasting {#lit}
\textcolor{blue}{
There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare. Areas such as forecasting call volume arrivals, ambulance demand and Emergency Department forecasting has received a significant attention. We refer interested readers to  some extensive reviews of the relevant literature} [@mingliterature2022;@gul2020review;@ibrahim2016modeling]. \textcolor{blue}{The time granularity considered by these studies spans from hourly to yearly across different parts of the healthcare. However, given
the focus of this study, we only discuss hourly ED forecasting.}

@@

Figure \@ref(fig:seasonplot-dofw) \textcolor{blue}{illustrates the daily subseries plot, with the x-axis representing the date and y-axis the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day of week (i.e. seven days) from Monday to Saturday. The blue line shows the average arrival for the given day. It is clear that ED arrivals on Monday are higher followed by Saturday. This indicates that there are
significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.}

@@

Probabilistic forecasts are evaluated following the principle of
*sharpness subject to calibration*, meaning that the sharper forecast is
prefered provided that it is calibrated. Mis-calibrated forecasts are
unsuitable for use in decision-making, so they should be excluded.
Calibration is evaluated visually in Figure \@ref(fig:quantile-bias),
which highlights a systematic negative bias across all probability
levels in many models, with only the truncated normal and $t$ family
GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks
exhibit negative quantile bias as they struggle to capture the long term trend of increasing arrivals. This could result in poor staffing decisions. This is because the empirical distribution of whole data fails to characterise how arrivals in ED may change over time.}

@@

\textcolor{blue}{The benefits of the proposed model and its managerial aspects include
the use of probabilistic forecast to inform decision making and provide
insights on the running time of each model to generate forecasts. Probabilistic forecasts contain all potential future outcomes and help
planners to achieve a more efficient decision making by not only allocate a probability to the most likely outcome but also to unlikely
outcomes such as extremely high or low arrivals in the Emergency Departments. The running time could be also an important aspects for managers depending on the frequency of generating forecasts. If the frequency of forecast generation is high, one may employ models such as
Poisson regression (Poisson 1). If the frequency is lower and running time is not a big concern, a model like GAMLSS (Ttr-2) should be used.}

\textcolor{blue}{Probabilistic forecasts can be considered as a useful tool for planners to manage the risk of under and over resource allocation, which consequently may reduce both costs and risks of service delivery in Emergency Departments.}

@@

\textcolor{blue}{
The dataset used in this study does not include the period of the COVID-19 pandemic. During COVID-19, the dynamic of ED has changed. That means any forecasting model that is used during such periods to forecast ED arrivals needs to be modified to reflect those changes for that specific period. However, we expect that if the COVID period is taken into account by the models (or interpolated), the models should perform similarly.}

>>>>

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average
(ARIMA) [@hyndman2021forecasting], Autoregressive Conditional
Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive
model [@lutkepohl2013vector], TBATS [@de2011forecasting] and Artificial
Neural Networks [@hyndman2021forecasting]. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{There is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric, i.e. having more staff than needed is costly, but
having less staff than required may lead to higher death rate}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient [@wright2006reexamining]. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently [@ramos2013probabilistic;@ROSTAMITABAR20221197]. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow _optimal_ decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{nother drawback of existing studies is that the datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to capture the inter-annual seasonality correctly and to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and zeros in some hours of the day, which brings additional challenges to traditional time series forecasting approaches. Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.}

@@

# Research background: hourly ED forecasting {#lit}
\textcolor{blue}{
There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare. Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention. We refer interested readers to some extensive reviews of the relevant literature by @mingliterature2022, @gul2020review and @ibrahim2016modeling. \textcolor{blue}{The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare. However, given the focus of this study, we only discuss hourly ED forecasting.}

@@

Figure \@ref(fig:seasonplot-dofw) \textcolor{blue}{ illustrates the daily subseries plot, with the x-axis representing the date and y-axis
the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day
of week from Monday to Saturday. The blue line shows the average arrival for the given day.
It is clear that ED arrivals on Mondays are higher than on other days. This is followed by
Saturday. This indicates that there are significantly more arrivals on Mondays and Saturdays
compared to the rest of the week. This might be due to the closure of General Practitioners
outpatient clinics over the weekends.}

@@

Probabilistic forecasts are evaluated following the principle of
*sharpness subject to calibration*, meaning that the sharper forecast is
prefered provided that it is calibrated. Mis-calibrated forecasts are
unsuitable for use in decision-making, so they should be excluded.
Calibration is evaluated visually in Figure \@ref(fig:quantile-bias),
which highlights a systematic negative bias across all probability
levels in many models, with only the truncated normal and $t$ family
GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks exhibit nega-
tive quantile bias as they struggle to capture the long term trend of increasing arrivals.
This could result in poor staffing decisions. This is because the empirical distribution of
whole data fails to characterise how arrivals in ED may change over time.}

@@

\textcolor{blue}{he main benefit of the applied models is the use of probabilistic forecasts to inform decision
making. Probabilistic forecasts contain all potential future outcomes and help planners to
achieve more efficient decisions by not only allocating a probability to the most likely outcome
but also considering less likely outcomes, including extremely high or extremely low arrivals
in the Emergency Departments. Based on these forecasts, hospitals can decide how many
nurses to have for each shift to make the work of the ED more efficient, e.g. to meet the
service level targets set by the National Health Service (NHS) in the United Kingdom.
Practically speaking, the running time can be an important aspect for managers depending
on the frequency of generating forecasts. If it is high, one may employ models that do not
require a lot of computational time, such as Poisson regression (Poisson 1), sacrificing the
accuracy of forecasts. If the frequency is lower and running time is not a big concern, a model
like GAMLSS (Ttr-2) should be used}

\textcolor{blue}{Probabilistic forecasts of daily ED arrivals can benefit ED process because it provides the timing and the magnitude of the unlikely scenarios with huge impact on the service delivery, which are fundamental for capacity planning. Probabilistic forecasts can be used to better manage risks of under and over resource allocation, which consequently can reduce both costs and risks for patients, staff and the service as a whole.}

@@

\textcolor{blue}{
The dataset used in this study does not include the period of the COVID-19 pandemic.
During COVID-19, the dynamics of ED arrivals has changed substantially. This means that
any forecasting model used for ED arrivals forecasting during that period would need to be
modified to reflect those changes for that specific period. One of the simplest modifications
would be to include a set of dummy variables, capturing different stages of the pandemic.
However, this is outside of the scope of this paper and can be considered as a direction for
future research.}

### 2022-12-29 paper/paper.tex
incorporating changes proposed by Ivan in the response letter

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average
(ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional
Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive
model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial
Neural Networks \citep{hyndman2021forecasting}. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient \citep{wright2006reexamining}. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently \citep{ramos2013probabilistic, ROSTAMITABAR20221197}. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow \emph{optimal} decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{Another drawback of existing studies is that the
datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
Finally, all previous publications
referenced in this paper are not fully reproducible as underlying data and code are not available.}

@@

\textcolor{blue}{
There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare. Areas such as forecasting call volume arrivals, ambulance demand and Emergency Department forecasting has received a significant attention. We refer interested readers to  some extensive reviews of the relevant literature} \citep{mingliterature2022, gul2020review, ibrahim2016modeling}. \textcolor{blue}{The time granularity considered by these studies spans from hourly to yearly across different parts of the healthcare. However, given
the focus of this study, we only discuss hourly ED forecasting.}

@@

Figure \ref{fig:seasonplot-dofw} \textcolor{blue}{illustrates the daily subseries plot, with the x-axis representing the date and y-axis the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day of week (i.e. seven days) from Monday to Saturday. The blue line shows the average arrival for the given day. It is clear that ED arrivals on Monday are higher followed by Saturday. This indicates that there are
significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.}

@@

The variables \emph{1--3} allow modelling the seasonal patterns on the
appropriate level of detail throughout the year, while \emph{4} covers the
changes in admittance due to calendar events. Finally, \emph{5} is needed to
capture the potential phenomenon of change in admittance after the
holiday (e.g.~people might try not to go to the hospital on Christmas
eve and thus go the next day). This model assumes that all these effects
are deterministic and do not change over time. Still, the exponentiation
in \eqref{eq:PoissonRegression} introduces an interaction effect between
dummy variables so that the 3 pm on Monday in January will be different
from 3 pm on Monday in July, although the parameters for the hour of day
and day of the week are fixed and do not change over time. We use the
\texttt{alm()} function from the \texttt{greybox} package \citep{Svetunkov2021Greybox} for
R \citep{RTeam2021} for the experiments and denote this model as ``Poisson
Regression''.

@@

Probabilistic forecasts are evaluated following the principle of
\emph{sharpness subject to calibration}, meaning that the sharper forecast is
prefered provided that it is calibrated. Mis-calibrated forecasts are
unsuitable for use in decision-making, so they should be excluded.
Calibration is evaluated visually in Figure \ref{fig:quantile-bias},
which highlights a systematic negative bias across all probability
levels in many models, with only the truncated normal and \(t\) family
GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks
exhibit negative quantile bias as they struggle to capture the long term trend of increasing arrivals. This could result in poor staffing decisions. This is because the empirical distribution of whole data fails to characterise how arrivals in ED may change over time.}

@@

\textcolor{blue}{
The dataset used in this study does not include the COVID-19 pandemic. During COVID-19, the dynamic of ED has changed. That means any forecasting model that is used during such periods to forecast ED arrivals needs to be modified to reflect those changes for that specific period. However, we expect that if the COVID period is taken into account by the models (or interpolated), the models should perform similarly.}

>>>>

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average
(ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional
Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive
model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial
Neural Networks \citep{hyndman2021forecasting}. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{There is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric, i.e. having more staff than needed is costly, but
having less staff than required may lead to higher death rate}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient \citep{wright2006reexamining}. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently \citep{ramos2013probabilistic, ROSTAMITABAR20221197}. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow \emph{optimal} decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{nother drawback of existing studies is that the datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to capture the inter-annual seasonality correctly and to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and zeros in some hours of the day, which brings additional challenges to traditional time series forecasting approaches. Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.}

@@

\textbackslash textcolor\{blue\}\{
There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare. Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention. We refer interested readers to some extensive reviews of the relevant literature by \citet{mingliterature2022}, \citet{gul2020review} and \citet{ibrahim2016modeling}. \textcolor{blue}{The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare. However, given the focus of this study, we only discuss hourly ED forecasting.}

@@

Figure \ref{fig:seasonplot-dofw} \textcolor{blue}{ illustrates the daily subseries plot, with the x-axis representing the date and y-axis
the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day
of week from Monday to Saturday. The blue line shows the average arrival for the given day.
It is clear that ED arrivals on Mondays are higher than on other days. This is followed by
Saturday. This indicates that there are significantly more arrivals on Mondays and Saturdays
compared to the rest of the week. This might be due to the closure of General Practitioners
outpatient clinics over the weekends.}

@@

The variables \emph{1--3} allow to model the seasonal patterns on the
appropriate level of detail throughout the year, while \emph{4} covers the
changes in admittance due to calendar events. Finally, \emph{5} is needed to
capture the potential phenomenon of change in admittance after the
holiday (e.g.~people might try not to go to the hospital on Christmas
eve and thus go the next day). This model assumes that all these effects
are deterministic and do not change over time. Still, the exponentiation
in \eqref{eq:PoissonRegression} introduces an interaction effect between
dummy variables so that the 3 pm on Monday in January will be different
from 3 pm on Monday in July, although the parameters for the hour of day
and day of the week are fixed and do not change over time. We use the
\texttt{alm()} function from the \texttt{greybox} package \citep{Svetunkov2021Greybox} for
R \citep{RTeam2021} for the experiments and denote this model as ``Poisson
Regression''.

@@

Probabilistic forecasts are evaluated following the principle of
\emph{sharpness subject to calibration}, meaning that the sharper forecast is
prefered provided that it is calibrated. Mis-calibrated forecasts are
unsuitable for use in decision-making, so they should be excluded.
Calibration is evaluated visually in Figure \ref{fig:quantile-bias},
which highlights a systematic negative bias across all probability
levels in many models, with only the truncated normal and \(t\) family
GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks exhibit nega-
tive quantile bias as they struggle to capture the long term trend of increasing arrivals.
This could result in poor staffing decisions. This is because the empirical distribution of
whole data fails to characterise how arrivals in ED may change over time.}

@@

\textcolor{blue}{
The dataset used in this study does not include the period of the COVID-19 pandemic.
During COVID-19, the dynamics of ED arrivals has changed substantially. This means that
any forecasting model used for ED arrivals forecasting during that period would need to be
modified to reflect those changes for that specific period. One of the simplest modifications
would be to include a set of dummy variables, capturing different stages of the pandemic.
However, this is outside of the scope of this paper and can be considered as a direction for
future research.}
